---
title: "LogFSM Tables (Output)"
author: "Ulf Kroehne"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LogFSM Tables (Output)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
   
`%>%` <- magrittr::`%>%`
```
  
##  Overview 

Schematic illustration of LogFSM input and output

```{r, echo=F, out.width="100%"}
 
p1 <-" digraph G {
          rankdir=LR; 
          node [shape=record, fontname = Arial];
          a [label = '<f0> Data | <f1> Syntax | (Parameters and Flags)', width = 3];
          b [label = '<f0> Augmented Data | <f1> Graph', width = 2.5];
          a:f0 -> b:f0 
          a:f1 -> b:f1 
           
          tab2 [label = '(aggregated)' shape=ellipse, color=gray]; 
          tab1 [label = 'for each machine i'  shape=ellipse, color=gray]; 
        
          b:f0 -> tab1:f0   
          b:f0 -> tab2:f0   
          c [label = '<f0> AugmentedLogDataTable', width = 3];
         
          d [label = '<f0> SequenceTable_i', width = 3];
          e [label = '<f0> TransitionFrequencyTable_i', width = 3];
          f [label = '<f0> NGramTable', width = 3];
          g [label = '<f0> StateEventFrequencyTable_i', width = 3];
          h [label = '<f0> TransitionFrequencyTotalsTable', width = 3];
          i [label = '<f0> VariableValueTable', width = 3];
          j [label = '<f0> UmlDotGraph [[i]]', width = 3];
          l [label = '<f0> StateFrequencyGraph [[i]]', width = 3];
          k [label = '<f0> StateSummaryTable_i', width = 3];
          
          subgraph cluster_0 {
              label='Input';
             a
          }
          subgraph cluster_1 {
              label='Output';
              c; b; d; e; f; g; h; i; j; l; tab1; tab2; k
          }
    
          tab1:f0 -> d:f0 
          tab1:f0 -> e:f0 
          tab1:f0 -> f:f0 
          tab1:f0 -> g:f0 
          tab1:f0 -> k:f0 
          
          tab2:f0 -> h:f0 
          tab2:f0 -> i:f0 
          b:f0 -> c:f0 
          b:f1 -> j:f0 
          b:f1 -> l:f0
          
      }" 
 
 DiagrammeR::grViz(p1)
 
```


Function ```LogFSM::RunFSMSyntax()``` Returns a List (e.g., ```out```)

- The following generic tables are created using all machines
 
```{r, eval=F, echo=T}
View(out$AugmentedLogDataTable)   
View(out$VariableValueTable)   
View(out$TransitionFrequencyTotalsTable)   
```
 
- Additional tables are created for each machine (e.g., machine $1$)

```{r, eval=F, echo=T}
View(out$StateEventFrequencyTable_1)  
View(out$SequenceTable_1)   
View(out$StateSummaryTable_1)
View(out$TransitionFrequencyTable_1)
View(out$NGramTable_1)
```
  
- DOT (GraphViz) syntax for plotting is created (e.g., machine $1$)

```{r, eval=F, echo=T}
DiagrammeR::grViz(out$UmlDotGraph[[1]])                 # input
DiagrammeR::grViz(out$TransitionFrequencyGraph[[1]])    # output
```
 
Step-by-step procedure for the creation of finite state machines for empirical data sets 
 
```{r, out.width='100%', echo=F}
 
p1 <-" digraph G {
          graph [ranksep='0.175', overlap = true, fontsize = 12];  
          node [shape=record, fontname = Arial];
          step1 [label = 'Start with Minimal FSM Syntax', width = 4];
          step2 [label = 'Run FSM With Selected Case', width = 4];
          step3 [label = 'Inspect Output Tables and Graphs', width = 4];
          check [label = 'Are the unexpeced \\nrejected events or \\nmissing state changes?', shape='ellipse'];
          step4 [label = 'Extend / Modify FSM Syntax', width = 4];
          step5 [label = 'Run FSM With All Case', width = 4];
          step6 [label = 'Extrat Indicator using R Syntax...', width = 4];
          step1 -> step2 -> step3  
          step4 -> step2  [label='Gradually increase  \\nthe number of cases \\n in order to \\n progressively \\n develop the FSM...', 
                           fontsize = 10, lp=30]; 
          check -> step4 [label='Yes']; 
          step3 -> check
          check -> step5  [label='No']; 
          step5 -> step6
          step6 ->step2  [label='If unexpected or \\nimplausible results \\noccur, consider these \\ncases separately...' , fontsize = 10]; 
      }" 
DiagrammeR::grViz(p1)
``` 
 
## AugmentedLogDataTable

```{r, eval=F, echo=T, tidy=F}
View(out$AugmentedLogDataTable)
```

Basic Structure of the ```AugmentedLogDataTable```

- Flat log data table for all processed test-takers

```{r, eval=T, echo=F, warning=F}

df1 <- data.frame(V1=c("ID01","ID01", "...","ID02", "..."),
                 V2=rep("...",5),
                 V3=rep("...",5),
                 V4=rep("...",5),
                 V5=rep("...",5),
                 V6=rep("...",5))

colnames(df1) <- c("PersonIdentifier","{Time}","Element","EventName",
                   "Event-Specific Data e(x)", "...")

df1 %>% kableExtra::kable() %>%  kableExtra::kable_styling(font_size = 12)
```
  
- *PersonIdentifier* is the identifier for persons (students, ...)
- *{Time}* is either ```TimeStamp``` (default) or ```RelativeTime```^[If the flag ```RELATIVETIME``` is set]
- *EventName* as the name of the log event (used as trigger if no attribute is specified)
- *Element* to identify parts of the assessment (e.g., item, unit, ...) 
 
Result of the Finite-State Machines

- For each defined finite-state machine^[e.g., for the first machine with *{Machine Index}* = 1] the following information is added
  
  - *State_Before_{Machine Index}*: The current state of the *FSM* with index *{Machine Index}* is added in this column
  
  - *State_After_{Machine Index}*: When for the given state a transition is defined, and the log data in a row match to a defined trigger, the *FSM* accepts the input and changes to a new state
  
  - *State_Result_{Machine Index}*: The result of processing a line of the input (Possible values are: ```EventAccepted```, ```EventNotDefinedAsTrigger```, and ```EventNotPermitted```)
  
- AugmentedLogDataTable can be used to retrace line by line the behavior of the finite-state machine 


Columns Added by LogFSM (Augmented Data)

- For each machine, the columns *State_Before_{Machine Index}, *State_After_{Machine Index} and *Result_{Machine Index} are added to the table (= the input is augmented)

```{r, eval=T, echo=F, warning=F}

df2 <- data.frame(V1=rep("...",3),
                 V2=rep("...",3),
                 V3=rep("...",3),
                 V4=rep("...",3),
                 V5=rep("...",3),
                 V6=rep("...",3),
                 V7=rep("...",3))

colnames(df2) <- c("...","TimeDifference", "StateBefore_1", 
                   "StateAfter_1", "Result_1","{Variables}", "...")

df2 %>% kableExtra::kable() %>%  kableExtra::kable_styling(font_size = 12)
```

- *TimeDifference*: The time difference to the previous event (computed either from relative times or absolute timestamps)

- *{Variables}*: A column for each defined variable is added
  - Variables are ```NA``` before the first value is assigned to a variable
  - Variable values are carried forward to the next event by default
  
## TransitionFrequencyTotalsTable 

```{r, eval=F, echo=T, tidy=F}
View(out$TransitionFrequencyTotalsTable)
```
 
- For all machines, the frequency for occurred transitions is aggregated

- Can be used to inspect the specification (are all expected transitions observed?)

- Structure of the ```TransitionFrequencyTotalsTable```:

```{r, eval=T, echo=F, warning=F}

df3 <- data.frame(V1=c("1","1", "...","1", "..."),
                 V2=c("Startstate","State1","...","State3","..."),
                 V3=c("State1","State2","...","Endstate","..."),
                 V4=c("100","...","...","...","...") )

colnames(df3) <- c("Machine","From","To","Frequency")

df3 %>% kableExtra::kable() %>%  kableExtra::kable_styling(font_size = 12)
 
```  

## TransitionFrequencyTable 

```{r, eval=F, echo=T, tidy=F}
View(out$TransitionFrequencyTable_1)
```

- Structure of the ```TransitionFrequencyTable_i```

   - All transitions with a frequency $> 0$ are listed, for each test-taker
   - Table is created for each machine separately ($i=1$ for the first machine)

```{r, eval=T, echo=F, warning=F}

df1 <- data.frame(V1=c("ID01","ID01", "..."),
                 V2=rep("...",3),
                 V3=rep("...",3),
                 V4=rep("...",3))

colnames(df1) <- c("PersonIdentifier","From","To", "Frequency")

df1 %>% kableExtra::kable() %>%  kableExtra::kable_styling(font_size = 12)
```

- Example usage to create an indicator 

```{r logfsm-output-transitionfrequencytable-02, eval=F, echo=T, tidy=F}
out$TransitionFrequencyTotalsTable %>% 
  filter(From == "Page7" & To == "Page8") %>% 
  select(Frequency)
```


## UmlDotGraph

Generic Visualization of the Decomposition 

- LogFSM creates / exports DOT language for visualization of state machines using the DOT^[Can plotted in R using the package DiagrammeR, see http://rich-iannone.github.io/DiagrammeR/] language
- Provided in the result list in an additional list ```UmlDotGraph``` (one diagram for each machine)
 
- Example  

```{r, eval=F, echo=T, warning=F}
out <- RunFSMSyntax(...)
library(DiagrammeR)
grViz(out$UmlDotGraph[[1]])
```
 
- Make sure to add the ```[[1]]```; otherwise an error will occur

## StateFrequencyGraph

Generic Visualization of Results 

- Visualizes data from the ```TransitionFrequencyTotalsTable```
- Nodes are the defined state (1:1 from the state machine)
- Transitions are reduced to the empirically observed transitions 
- Weight of the path corresponds to the proportion (frequency of transition divided by the total number of transitions, **within states**)
- Rule: Sum of all outgoing path from one node is $1.0$
 
- Example  

```{r, eval=F, echo=T, warning=F}
out <- RunFSMSyntax(...)
library(DiagrammeR)
grViz(out$TransitionFrequencyGraph[[1]])
```

## StateEventFrequencyTable 

```{r, eval=F, echo=T, tidy=F}
View(out$StateEventFrequencyTable_1)
```

- Frequencies of all events (column ```Name```) within each state are counted per test-taker 

```{r, eval=T, echo=F, warning=F}

df1 <- data.frame(V1=c("ID01","..."),
                 V2=rep("...",2),
                 V3=rep("...",2),
                 V4=rep("...",2),
                 V5=rep("...",2))

colnames(df1) <- c("PersonIdentifier","State","Name", "Result", "Frequency")

df1 %>% kableExtra::kable() %>%  kableExtra::kable_styling(font_size = 12)
```

- Result refers to the result of the FSM (one table for each machine)
 
- Typical Usage 

    - Explore log data (occurrence of contextualized events) 
    - Create indicators using activities (i.e., log information that has no duration or where the duration cannot be reconstructed from the log data) 
 
## SequenceTable

```{r, eval=F, echo=T, tidy=F}
View(out$SequenceTable_1)
```

- Table is the core output of LogFSM, one table for each machine
- One line per state (i.e., the reconstructed sequence is the combination of all lines)

```{r, eval=T, echo=F, warning=F}

df1 <- data.frame(V1=c("ID01","ID01","..."),
                 V2=c("State1","State2","..."),
                 V3=rep("...",3))

colnames(df1) <- c("PersonIdentifier","State","Duration")

df1 %>% kableExtra::kable() %>%  kableExtra::kable_styling(font_size = 12)
```

- Entries in the table are ordered by time (=**reconstructed sequence**)

Typical Usage

- Querying whether the reconstructed sequence of a test-taker contains a particular state (e.g., visited a page, showed a behavior, ...)
- Concatenate rows to extract complete sequence 

##  *N-Grams* / NGramTable

Definition of *N-Grams*

- N-Gram is a contiguous sequence of n items from a given sample of text or speech (i.e., alphabet)
  - Unigram / 1-gram: A; B; C
  - Bigram / 2-gram: A-B; A-C; B-C; ...
  - Trigram / 3-gram: A-B-C; A-C-B; B-A-C; B-C-A; ...
  - ...

\vspace{-0.25cm}  
- Concept of n-grams is used to analyze log file data [e.g., @HeIdentifyingFeatureSequences2015;@HaoTamingLogFiles2016]

- LogFSM generalize^[Generalization not yet satisfactory, because states can be identified from events only under special circumstances.] the idea from *Event n-grams* to *State n-grams* (using states as the unit / alphabet for the n-grams)

  - Each visited state is considered as *unigram*
  - Transitions (i.e., ordered pairs of states) are considered as *bigram*
 
NGramTable 

```{r, eval=F, echo=T, tidy=F}
View(out$NGramTable_1)
```
 
- All n-grams of states are counted^[The max size for n-grams can be reduced using the parameter ```NGRAMMAXSIZE```] in  ```NGramTable_i```

```{r, eval=T, echo=F, warning=F}

df1 <- data.frame(V1=c("ID01","ID01", "..."),
                 V2=rep("...",3),
                 V3=rep("...",3),
                 V4=rep("...",3),
                 V4=rep("...",3))

colnames(df1) <- c("PersonIdentifier","Size","Sequence", "Frequency", "TotalTime")

df1 %>% kableExtra::kable() %>%  kableExtra::kable_styling(font_size = 12)
```

- Column ```Size``` refers to the size of the n-gram (1=unigram, 2=bigram, 3=trigram, ...)
- Column ```Sequence``` contains the states, separated by ```;```
- Column ```Frequency``` counts how often the particular n-gram was observed for the test-taker
- Column ```TotalTime``` is the cumulated time for this n-gram

## StateSummaryTable  

```{r, eval=F, echo=T, tidy=F}
View(out$StateSummaryTable_1)
```

- For each test-taker, all visited states are listed (one table per machine)

```{r, eval=T, echo=F, warning=F}

df1 <- data.frame(V1=c("ID01","ID01", "..."),
                 V2=rep("...",3),
                 V3=rep("...",3),
                 V4=rep("...",3),
                 V5=rep("...",3),
                 V6=rep("...",3),
                 V7=rep("...",3))

colnames(df1) <- c("PersonIdentifier","State","TotalTime", "Frequency", "ShortesttVisit", "LongestVisit", "AverageVisit")

df1 %>% kableExtra::kable() %>%  kableExtra::kable_styling(font_size = 12)
```

- Time information is provided as *TotalTime*, *ShortesVisit*, *LongestVisit*, and *AverageVisit*
- Frequency refers to the number of state visits (states with a frequency of $0$ are not included)
 
Typical Usage for Indicators 

- Filter relevant states (frequency, total time or both) and add $0$ or $NA$ for persons never visited the state(s) of interest

## VariableValueTable
 
```{r, eval=F, echo=T, tidy=F}
View(out$VariableValueTable)
```
 
- For each of the variables (defined by using them in operators), the last value, and the number of changes is summarized

```{r, eval=T, echo=F, warning=F}

df1 <- data.frame(V1=c("ID01","..."),
                 V2=rep("...",2),
                 V3=rep("...",2),
                 V4=rep("...",2))

colnames(df1) <- c("PersonIdentifier","Variable","Value", "NumberOfChanges")

df1 %>% kableExtra::kable() %>%  kableExtra::kable_styling(font_size = 12)
```

- Initializing the (default) value of a variable counts as variable change, i.e., the number of (real) value changes is typically 
- This table contains variables from all machines (i.e., only one table is created)
  
Typical Usage 

- Count of transitions, events, ... (i.e., allow to create counting indicators directly while running the state machine)
   
