---
title: "LogFSM: Transform Raw Log Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Transform}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette is aimed at data provides who want to provide log data in the universal log format. LogFSM provides a function to transform raw log data from selected supported platforms. 

## Transform Raw Log Data to Universal Log Format with LogFSM

LogFSM provides support to transform log data to the universal log  format currently for the following assessment platforms (possible values of the argument 'inputformat'):

* CBA ItemBuilder (Version <= 8.12), RAP Generation, Execution Environment (EE): ```eeibraprawv01a```
* CBA ItemBuilder (Version >= 8.12), REACT Generation, 'ItemBuilder Static Delivery' (IBSD): ```ibsdraw01a```
* NEPS-TBT-Moduls (Data gathered in the National Educational Panel Study, NEPS): ```nepsrawv01a```
 
### General Workflow: Raw Data to Universal Log Format

Input:

* Folder with raw log data 
* File name for output zip
* Input format (of a supported platform)
* Optional: Concordance table
* Optional: Codebook dictionary

Output:

* ZIP archiv with log files in the universal log data format
* Optional: Excel document as codebook
* Optional: Template of a concordance table

Sipmle Example: 

```{r, eval = FALSE}
myinfolder <- ""    # Absolute file path to the directory containing ZIP archives 
                    # or XML files exported from the CBA ItemBuilder Execution Environment 
mystatazip <- ""    # Absolute file path for a ZIP archive that will be created containing the
                    # log data in the universal log format
mycodebook <- ""    # Absolute file path for the XLSX document that will be created containg the
                    # codebook
mydictionary <- ""  # Absolute file path for a XLSX document with event-related platform-specific
                    # documentation that will be used as dictionary while creating the codebook

TransformToUniversalLogFormat(input = myinfolder
                              stata = mystatazip, 
                              inputformat = "eeibraprawv01a", 
                              dictionary = mydictionary, 
                              codebook = mycodebook)
```


### Specific Comments: NEPS-TBT-Moduls

Input:

* Input folders are expected to contain the extracted zip archives of individual assessments (raw data archives) created with NEPS 'Testanwendung'.

### Specific Comments: IB EE / RAP

Input:

* Input folders are expected to contain xml files (or zip archives with xml files) exported from the Execution Environment.

### Specific Comments: IBSD / REACT

Input:

* Input folders are expected to contain json files exported from the ItemBuilder static delivery environment.

## Documentation (Codebook)

### Generated Codebook

Codebooks contain the following sheets:

* ```MetaData```: General metadata about the log data.
* ```Head```: Variable labels for columns in all tables.
* ```Events```: Description of all events present in the log data.
* ```Attributes```: Description of all attributes (for the occured events).

For all entries in the sheet ```Head``` the column ```Identifies``` is provided. This column indicates which information is identified by the entries in this column. Possible values are: 

* ```line```: The values in this column can be used to identify lines. 
* ```target-person```: The values in this column identifiy the target-persons. 
* ```instrument-part```: The values in this column identify the administered part of the assessment instrument. 
* ```assessment-time```: The values in this column identify the assessment time. 
* ```event```: The values in this column identify events.
* ```table```: The values in this column identify tables.
* (empty or ```-```): This column is not an identifier. 


The table ```Attributes``` lists for each table which additional event-specific columns are included. For each column, you can specify whether it can contain anonymization-relevant information (column ``Anonymity``).  Possible values are: 

* ```instrument-part```: Entries in this column can contain information from the items used.
* ```target-person```: Entries in this column may contain information related to a person identifier. 
* ```paradata```: Entries in this column can contain paradata. 
* ```text response```: Entries in this column can contain entered text responses. 
* (empty or ```-```): The values in this column cannot contain information relevant to anonymization. 

The table ``Attributes``` also allows adding additional comments regarding potential steps necessary to purify the log data in the column ```Purification```. Moreover, ```Conditions``` can be described in the column of the same name, if attributes only under certain circumstances attribute values can contain anonymity-relevant information. 

### Dictionary (Codebook Template)

An XLSX document can be used as a template (dictionary) for creating a codeobk. 

Dictionary files can contain the following sheets:

* ```MetaData```: General metadata about the log data
* ```Head```: Variable labels for columns in all tables
* ```Events```: Description of all events
* ```Attributes```: Description of all event attributes

Language variants can be defined optionally. For this purpose, the selected columns can be defined several times in the following tables (column names with suffix _LANGUAGE, e.g., AttributeValue_ENG and AttributeValue_DE): 

* ```MetaData```: AttributeValue_LANGUAGE
* ```Head```: VariableLable_LANGUAGE
* ```Events```: EventDescription_LANGUAGE
* ```Attributes```: Description_LANGUAGE



The following example shows how a codebook dictionary can be created with the R package openxlsx 

```{r, eval = FALSE}
# create data for sheet 'metadata'

m <- rbind(c("StudyName", "Name of the study" ,"Studienname"),
           c("TestPlatform", "Software version","Version der Software"))

colnames(m) <- c("Attribute","AttributeValue_ENG","AttributeValue_DE")
```

```{r, eval = FALSE}
# create data for sheet 'head'

h <- rbind(
  c("ID", "ID for this line (counter over all cases and events)", 
          "ID für diese Zeile (Zähler über alle Fälle und Ereignisse)", 
    "line"),
  c("PersonIdentifier", "ID of the person which triggered the event data in this row",
                        "ID der Person, die das Log-Ereignis in dieser Zeile ausgelöst hat", 
    "target-person"),
  c("Element", "Item or page name (source of the event data in this row)",
               "Item oder Seitennahme (Quelle der Eventdaten in dieser Zeile)", 
    "instrument-part"),
  c("TimeStamp", "Time stamp for the event data in this line",
                 "Zeitstempel für die Ereignisdaten in dieser Zeile", 
    "assessment-time"),
  c("RelativeTime", "Relative time for the log event (milliseconds relative to the start)",
                    "Relativzeit für das Log-Ereignis (Millisekunden relativ zum Start)", 
    "-"),
  c("EventID", "ID for the current event (starts with 0 within the current 'Path')",
               "ID für das aktuelle Ereignis (beginnt mit 0 innerhalb des aktuellen 'Path')", 
    "event"),
  c("ParentEventID", "ID of the parent event (used for the nested data structures of an event)",
                     "ID des übergeordneten Ereignisses (wird für die geschachtelten Datenstrukturen eines Ereignisses verwendet)",
    "event"),
  c("Path", "Hierarchy of the nested data structure",
            "Hierarchie der geschachtelten Datenstruktur", 
    "table"),
  c("ParentPath", "Hierarchy of the parent element (empty for the elements at the root level)",
                  "Hierarchie des Elternelements (leer für die Elemente auf der Wurzelebene)", 
    "table"),
  c("EventName", "Name of the log event / data in this line",
                 "Name des Logevents in dieser Zeile", 
    "-"))

colnames(h) <- c("Column","VariableLable_ENG","VariableLable_DE","Identifies")
```

```{r, eval = FALSE}
# create data for sheet 'events'

e <- rbind(
  c("ButtonLogEntry",
    "Log.ButtonLogEntry",
    "Jeder Interaktion eines Testteilnehmers mit einer Schaltfläche führt zum Speichern eines Log-Ereignisses vom Typ ButtonLogEntry. Neben dem Zeitstempel wird im Attribut a_id gespeichert, welche Schaltfläche gedrückt wurde. Die Namen der Schaltflächen sind als so-genannte user-defined ID innerhalb des Items festgelegt. ",
    "Each interaction of a test participant with a button results in the saving of a log event of the type ButtonLogEntry. In addition to the timestamp, the a_id attribute stores which button was pressed. The names of the buttons are defined as a so-called user-defined ID within the item. ") # ...
)  

colnames(e) <- c("EventName","Table","EventDescription_DE","EventDescription_ENG")
```

```{r, eval = FALSE}
# create data for sheet 'attributes'

a <- rbind(
  c("Log.CBAItemLogEntry",
    "a_webClientUserAgent","-",
    "User Agent String des Browsers.",
    "User Agent String of the browser.",
    "paradata ","-"),
  c("Log.ButtonLogEntry",
    "a_id","-",
    "User-Defined ID der Schaltfläche, welche gedrückt wurde. Die Namen der Schaltflächen sind als so-genannte user-defined ID innerhalb des Items festgelegt. ",
    "User-Defined ID of the button that was pressed. The names of the buttons are defined as a so-called user-defined ID within the item. ",
    "-","-")# ...
)

colnames(a)<- c("Table","Column","Condition","Description_DE","Description_ENG", "Anonymity", "Purification")

```


```{r, eval = FALSE}

# combine datasets and write as xlsx file

b <- createWorkbook(f)

addWorksheet(b, "MetaData")
writeData(b, sheet = 1, m)

addWorksheet(b, "Head")
writeData(b, sheet = 2, h)

addWorksheet(b, "Events")
writeData(b, sheet = 3, e)
 
addWorksheet(b, "Attributes")
writeData(b, sheet = 4, a)

saveWorkbook(b, file = "dictionary.xlsx", overwrite = TRUE)
```


## Concordance Table 

If a file name is specified for a concordance table, and the file does not exist, then the TransformToUniversalLogFormat function creates a template file. This file contains all PersonIdentifiers and can be edited / modified for a second call of the function. 

### Change PersonIdentifier

A concordance table in XLSX, CSV or Stata format must have two columns:

* ```PersonIdentifierOld```:  Original ID
* ```PersonIdentifierNew```:  New ID

When creating log data in Universal Log Format, the original ID is exchanged for the new ID. 

### Filter Test-Takers

If an existing XLSX, CSV or Stata file is transferred as a concordance table, then only those cases are transferred to the Universal Log File for which they are contained in the concordance table. 

Cases that are not included in the concordance table are filtered (that is, ignored). 

## Transform Raw Log Data to Universal Log Format using Commmand Line Tool

In order to be able to carry out a conversion outside of R, a commandLine tool is provided to accompany the conversion. This allows access to the options available via the LogFSM function ```TransformToUniversalLogFormat``` from the command line. 

Arguments (Input): 

* ```-i``` / ```-input```: Input folder to be processed.
* ```-r``` / ```-rawformat```: Raw format (either ```eeibraprawv01a```, ```ibsdraw01a``` or ```nepsrawv01a```). 
* ```-l``` / ```-logversion```: Version information about the raw data.

Arguments (Output): 

* ```-s``` / ```-stata```: Output file name for the generated universal log format, type Stata (i.e., absolute path to the zip file containing log data as Stata files, one file for each event type).")
* ```-z``` / ```-zcsv```: Output file name for the generated universal log format, type zip-compressed CSV (i.e., absolute path to the zip file containing log data as CSV files, one file for each event type).
* ```-x``` / ```-xslx```: Output file name for the generated universal log format, type XLSX (i.e., absolute path to XLSX file containing log data, one sheet for each event type).

Filter:
 
* ```-m``` / ```-mask```: File filter mask. Only files that match the specified mask will be used (e.g., *.jsonl).
* ```-e``` / ```-excludedelements```: Element names (i.e., items, units or tasks), that should be ignored.
* ```-t``` / ```-table```: Concordance table file name. 

Codebook:     
 
* ```-r``` / ```-codebook```: File name for the XLSX file created as codebook. 
* ```-r``` / ```-dictionary```: Dictionary file for the creation of an integrated codebook.

Misc:

* ```-f``` / ```-flags```: Optional flags as documented for the specific data formats.
* ```-a``` / ```-arguments```: Additional arguments (URL style, i.e., ?name1=value&name2=value).

* ```-v``` / ```-verbose```: Request verbose output messages.

Example call:

```TransformToUniversalLogFormat.exe --i PATH\in\ -s PATH\out\data_dta.zip```

To request help information, call: 

```TransformToUniversalLogFormat.exe --help```

